The ChEMBL database as Linked Open Data

### 2012-04-15 article.tex
edit the abstract, introduction and methods sections

\begin{abstract}
Making data available as RDF has the advantages that integration with other web resources brings:
we can link out to related data, others can link to this data, it makes the data machine readable,
is easy to extend with additional information,  and enables a representation of the data in common
vocabularies. This combination provides a clear, open interface.
This paper describes the continued conversion of data from the ChEMBL database into RDF triples.
This updated version of ChEMBL-RDF now uses recently introduced, such as the CHEMINF and CiTO ontologies,
exposes more information from the database, and is now available as dereferencable, linked data.
To demonstrate the new features, we here present a few new use cases, showing integration with
other web resources using semantic web technologies.
\end{abstract}

@@

Despite the data deluge, finding new, unique, but significant patterns explaining biological
phenomena not yet understood is limited by the availability of data. Such data analysis is
beyond the scope of single data sets; it requires data integration from many sources, for
example, systems biology integrating micro-array differential expression to biological
pathways~\cite{}. Another prominent example is drug discovery where a new unique chemical entity is
searched with just the right biological properties. This too requires linking many
data sets~\cite{Samwald2011,OpenPHACTS}.

ChEMBL contains biological activities for over a million chemical entities, extracted from
literature, providing a unique resource for drug discovers~\cite{Gaulton2012,Warr2009}.
It is updated on a fairly frequent basis as the existing data is further curated and new data is added. Originally, this
data was available for download and via a web interface. The former requires setting up
the data in a relational database, while the latter limits the machine access to the data.
Two independent teams created RDF triples for the ChEMBL data, resulting in exposure
via Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}.

This paper presents the history of the ChEMBL-RDF data set, details of the latest structures
and ontologies used for the RDF version of ChEMBL 13, and a few new example use cases,
exemplifying how the data set can be linked to other data sources to further support
research in the life sciences.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
RDF triples were created from a local MySQL database to which the SQL data dump of ChEMBL was
inserted, using a PHP script, available from the source code hosting service
GitHub~\citep{ChEMBLRDFGitHub}, a process outlined earlier in a best practices note by
the international Health Care and Life Sciences XXX Working Group~\cite{Marshall2012}.

During the life time of this project, the ontologies used in ChEMBL-RDF have changed,
initially using a custom, ad-hoc ontology, but increasingly with community-proposed
ontologies, making the RDF more interoperable. The ChEMBL-RDF 13 dataset uses ontology
standards such as the Bibliography Ontology for Citation Typing Ontology for literature
references, and domain ontologies like the Protein Ontology and the Chemical Information
Ontology. Throughout this paper, various prefixes are used to simplify the RDF output and they are outlined
in Table~\ref{namespaces}.

@@

To expose that ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable using the Kasabi platform~\cite{kasabi}. This platform also provides a SPARQL
end point, but requires a user account.

An initial custom vocabulary has been defined, reflecting the concepts and links found in the
data structure of the ChEMBL database. This vocabulary has only partially been converted into
an OWL ontology, and is being phased out and replaced by common ontologies. This has not happened
for all data yet, and terms from the ChEMBL-RDF namespace (\textit{http://rdf.farmbio.uu.se/chembl/onto/\#},
prefix: \textit{chembl}) are still in use.

As a method to further test the access and possibilities of this Linked Data version of
ChEMBL, a few new uses cases have been developed for this paper.

>>>>

\begin{abstract}
Making data available as Linked Data using RDF is advantageous because it promotes integration with other web resources.
RDF documents can natively link to related data, and others can link back using URIs. RDF makes the data machine-readable,
using extensible vocabularies for additional information. This combination provides a new, open, interface for scientists to use ChEMBL.
This paper describes the continued conversion of data from the ChEMBL database into RDF triples.
This updated version of ChEMBL-RDF now uses recently introduced ontologies, including CHEMINF and CiTO,
exposes more information from the database, and is now available as dereferencable, Linked Data.
To demonstrate the new features, we present new use cases showing the benefits of integration with
other web resources using semantic web technologies.
\end{abstract}

@@

The current scientific data deluge, in which datasets are growing faster than scientists can analyse and 
curate them, provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,OpenPHACTS}.

ChEMBL contains descriptions for biological activities involving over a million chemical 
entities, extracted from literature. This provides a unique resource for drug researchers~\cite{Gaulton2012,Warr2009}.
It is updated on a fairly frequent basis as the existing data is further curated and new data is added. 
The ChEMBL dataset is available for download and can be browsed using a web interface. 
The former requires scientists to import the data into a relational database, while the 
latter limits the machine access to the data. Two independent teams have previously mapped 
the ChEMBL dataset to RDF, Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}.

This paper presents the current state of the ChEMBL-RDF dataset, details of the latest structures
and ontologies used to map version 13 of ChEMBL to RDF, along with new example use cases, showing 
links to other datasets using their RDF URIs to further support research in the life sciences.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL datadump provided by EMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF files using a set of PHP scripts. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\citep{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences XXX Working Group~\cite{Marshall2012}.

In the previous RDF conversion the RDF triples in ChEMBL-RDF were represented using properties and c
lasses from a custom, ad-hoc ontology~\cite{Willighagen2011}. The current version used community-proposed
ontologies, making the RDF more interoperable. The ChEMBL-RDF 13 dataset uses ontology
standards such as the Bibliography Ontology~\cite{Giasson2011} and the Citation Typing Ontology~\cite{Hastings2011} for literature
references, and domain ontologies like the Protein Ontology\cite{Sidhu2006} and the Chemical Information
Ontology\cite{Hastings2011}. Throughout this paper, various shorthand prefixes are used to denote different ontology 
namespaces to simplify the RDF examples. These prefixes are outlined in Table~\ref{namespaces}.

@@

To expose that ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable using the Kasabi platform~\cite{kasabi}. The Kasabi hosting provides a SPARQL
end point which requires a user account. A user account can be free for casual usage, or paid for consistent usage.

Although many parts of the CHEMBL-RDF dataset use community based ontologies, some 
terms from the previous ontology are still used. These terms were created under the 
URI namespace \textit{http://rdf.farmbio.uu.se/chembl/onto/\#}, and are referenced here using the
prefix: \textit{chembl}.

%As a method to further test the access and possibilities of this Linked Data version of
%ChEMBL,  new uses cases have been developed for this paper.
